Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-02-09 paper/manuscript.tex
Some edits to introduction

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g. binding affinities. 
These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment. 
Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
Alchemical calculations can be used in a variety of scenarios, such as: 
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{A});
\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
\end{itemize}
%

Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}. %
More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 
This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
%

As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This Best Practice guide provides current recommendations and tips for users of all experience.  
%
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule across a high barrier. \textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{D}: Effect of mutations of protein or host residues on free energies of binding. \textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g. protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
    \label{fig:fig_what_is_alchemy}
    }
\end{figure}

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
What can be expected from this Best Practice guide is a good starting point for new practitioners and a reference for experienced practitioners. For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations meet standard alchemical simulation and analysis practices. It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts. Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}. While reading this Best Practice document, it is important to bare in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead the guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical free energy calculation? 
    \item How do I choose the right free energy protocol and run it? 
    \item How do I analyze my simulations accurately? 
    \item What software tools are available to perform alchemical free energy calculations? 
\end{itemize}
%

Some other background information may be needed depending on the nature of the alchemical project. For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations. This will require some basic familiarity on how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}. 
%
As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.

@@

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate, aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item complex association stoichiometry (beyond simple 1:1 association).
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\end{itemize}
%

For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    \textit{CPU} -- Central Processing Unit\\
    \textit{BAR} -- Bennett Acceptance Ratio\\
    \textit{FEP} -- Free Energy Perturbation\\
    \textit{GPCR} -- G-Protein Coupled Receptor\\
    \textit{GPU} -- Graphics Processing Unit\\
    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} -- Maximum Common Substructure\\
    \textit{MD} -- Molecular Dynamics\\
    \textit{RMSE} -- Root Mean Square Error\\
    \textit{MUE} -- Mean unsigned error\\
    \textit{SAR} -- Structure-Activity Relationships\\
    \textit{TI} -- Thermodynamic Integration
\end{tcolorbox}

@@

\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
    \emph{ligand} - small molecule binding to a larger host molecule such as a protein (receptor).\\ 
    \emph{lead optimization} - process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
    \emph{docking} -- Method of generate simulation starting structures for ligand receptor systems, when no crystal structure is available\\
    \emph{thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
    \emph{alchemical} -- states or simulations that contain non-physical intermediates \\
\end{tcolorbox}

>>>>

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities. 
These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
Alchemical calculations can be used in a variety of scenarios, such as: 
\begin{itemize}
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{A});
\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
\end{itemize}
%
Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
\end{itemize}
%

After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate the free energy for the transformation of interest.
Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}. %
More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 
This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
%

As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon2013routine} and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform. 
In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
This Best Practices guide provides current recommendations and tips for users of all experience.  
%
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule across a high barrier. \textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{D}: Effect of mutations of protein or host residues on free energies of binding. \textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g. protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
    \label{fig:fig_what_is_alchemy}
    }
\end{figure}

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
What can be expected from this Best Practices guide is a good starting point for new practitioners and a reference for experienced practitioners. 
For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. 
Where the best practices are currently unsettled, we highlight areas where further research is needed to identify an unambiguous recommendation.
It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts. 
Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead the guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
    \item How do I select an appropriate alchemical protocol? 
    \item What software tools are available to perform alchemical calculations? 
    \item How should I analyze my data and report uncertainties? 
\end{itemize}
%

Some other background information may be required depending on the nature of the alchemical project. 
For example, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations. 
This will require some basic familiarity with how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}. 
%
As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.

@@

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate, aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item complex association stoichiometry (beyond simple 1:1 association).
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\end{itemize}
%

For convenience, we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    {\bf CPU} -- Central Processing Unit\\
    {\bf BAR} -- Bennett Acceptance Ratio\\
    {\bf FEP} -- Free Energy Perturbation\\
    {\bf GPCR} -- G-Protein Coupled Receptor\\
    {\bf GPU} -- Graphics Processing Unit\\
    {\bf MBAR} -- Multistate Bennett Acceptance Ratio\\
    {\bf MCSS} -- Maximum Common Substructure\\
    {\bf MD} -- Molecular Dynamics\\
    {\bf RMSE} -- Root Mean Square Error\\
    {\bf MUE} -- Mean unsigned error\\
    {\bf SAR} -- Structure-Activity Relationships\\
    {\bf TI} -- Thermodynamic Integration
\end{tcolorbox}

@@

\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
    {\bf ligand} -- small molecule binding to a larger host molecule such as a protein (receptor).\\ 
    {\bf lead optimization} -- process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
    {\bf docking} -- Method of generate simulation starting structures for ligand receptor systems, when no crystal structure is available\\
    {\bf thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
    {\bf alchemical} -- states or simulations that contain non-physical intermediates \\
\end{tcolorbox}
